The evolution of gene expression in the term placenta of viviparous mammals

### 2015-12-15 manuscript/manuscript.Rnw
use species aliases

One symptom
of this maternal-fetal conflict is its potential affect on human
health \cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. 
In addition to having an invasive
hemochorial placenta, humans possess a deeply invasive trophoblast
layer that can penetrate the myometrium of the uterus 
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
 Defects in placental growth are known to be
associated with the devastating condition of preeclampsia, which can
cause hypertension, proteinuria, and has the potential to lead to
death of the mother and/or infant \DLA{(citation)}. Although cases have been
documented in chimpanzees and gorillas \DLA{(citation)}, preeclampsia has a
far higher incidence in human pregnancies with as much as 4\% of
pregnancies being affected in the United States \DLA{(citation)}.
Chimpanzees and gorillas also have deeply invasive trophoblasts,
although the progression of invasion in great apes may be slower than
humans \cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however, large-scale expression
studies of great ape placentas have not been conducted. The increasing
affordability of next-generation sequencing technology has enabled the
large-scale study of specific tissues from a wide array of species to
address broader phylogenetic questions in the evolution of gene
expression \cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use Illumina RNA-Seq
technology to sequence and quantify the mRNA environment of six
mammalian placentas at term, including that of the human, bonobo (the
sister species to the chimpanzee), dog, cow, African elephant, and
opossum, representing marsupials as well as three major branches of
eutherian mammals (Euarchontoglires, Laurasiatheria, Afrotheria).
These placentas also represent a wide array of placental morphologies
as well as traits such as number and size of offspring and length of
gestation \DLA{(Table 1)}. We use this data, along with existing RNA-Seq
data from mouse \cite{Wang.ea2011:surveyfornovelimprinted}, to address the major differences
between expression in marsupial and eutherian mammals at term,
identify the evolution of expression in important placental genes such
as hormones and galectins, identify novel lineage-specific expression
patterns, and quantify human specific differences in gene expression
that may be related to obstetrical disorders such as preeclampsia.

@@

Fetal tissue was collected from the placenta of six mammalian species
(\textit{Homo sapiens} [human], \textit{Pan paniscus} [bonobo], \textit{Canis familiaris}
[domestic dog], \textit{Bos taurus} [domestic cow], \textit{Loxodonta africana} [African
savanna elephant], \textit{Monodelphis domestica} [gray short-tailed
opossum], \textit{Nannospalax gallili} [?], \textit{Spalax carmelli} [?]).
Information on specimen ID, place and time of collection,
and collector is shown for each sample in \DLA{Supplemental Table 1}. In the
case of \textit{Pan}, \textit{Bos}, and \textit{Loxodonta}, the placenta was collected at term
upon expulsion from the mother following delivery. Placental tissue
from \textit{Homo sapiens} was collected upon delivery via a caesarian section.
Near-term placental tissue from \textit{Canis familiaris} was collected upon
the death of the mother from a car accident. Placental tissue from
\textit{Monodelphis domestica} was sampled at fetal stage 32, roughly one day
prior to birth. Only the fetal portion of the placenta was sampled at
the maternal-fetal interface (i.e. villous portion for \textit{Homo}, \textit{Pan},
\textit{Canis}, and \textit{Loxodonta}; cotyledon for \textit{Bos}; diffuse placenta for
\textit{Monodelphis}). All tissues were collected and stored in RNAlater (Life
Technologies, Carlsbad, CA) until RNA isolation could be performed.
Tissues were homogenized in Trizol, and total RNA was isolated
following the manufacturer’s instructions. RNA was further purified
using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction with the
RNase-Free DNase Set (Qiagen, Valencia, CA), following the
manufacturer’s protocol to remove any contaminating DNA from the
sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
Clara, CA) to confirm concentration and quality of the isolation. RNA
quality was assessed using A260/A280 and A260/A230 absorbance values.

@@

For more complete comparisons additional sequences of placenta in
\textit{Mus musculus} \cite{Wang.ea2011:surveyfornovelimprinted},
\textit{Homo sapiens}
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \textit{Equus
  caballus} \DLA{citation}, \textit{Ovis aries} \DLA{citation}, and
\textit{Sus scrofa} \DLA{citation} were obtained from the \ac{SRA}
(Supplemental \cref{tab:srr_numbers}).

@@

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (Reference of ---
in \cref{tab:ref_genome_aligned_species}) were assembled into a
reference using Trinity v2014-07-17
\cite{Haas.ea2013:Denovotranscriptsequence} and annotated using
Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
the \textit{Homo sapiens} or \textit{Mus musculus} transcriptome
(\cref{tab:ref_genome_aligned_species}).
% 
All transcriptomes were then aligned to the appropriate reference
genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{Pan troglodytes} was used as
a reference for \textit{Pan paniscus}. The separation between the two
recognized Pan species is generally perceived to be $\approx 1$ Mya
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

@@

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using topGO
\cite{topGO:2.20.0}. Gene Ontology clusters were considered
significantly enriched if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure\cite{Benjamini1995:bh_procedure} on from the
$p$ values obtained from the Fisher Exact Test were less than or equal
to 0.05. Enrichment tests were considered with two gene lists per
transcriptome: annotated genes with expression levels of the top
5\% of the frequency distribution of \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris});
the top 100 genes with the highest \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
Due to the paucity of annotated gene ontology terms for some species,
gene lists were compared using the \textit{Homo sapiens} Gene Ontology
using the 1:1 \textit{Homo sapiens} ortholog.

@@

All 1:1 orthologues present among \textit{D. novemcinctus}, \textit{P.
  troglodytes}, \textit{B. taurus}, \textit{C. familiaris}, \textit{L.
  africana}, \textit{M. musculus}, \textit{H. sapiens}, and \textit{M.
  domestica} were determined using Ensembl v.80 gene trees via
Biomart\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the
biomaRt R package \DLA{Cite the biomaRt package}. A total of
\Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were
expressed in at least one species. Of these,
\Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  10,length(unique(human_name))]} 1:1 orthologous genes with
\ac{FPKM} $\ge 10$ were retained for future analyses of gene
expression evolution described below.

>>>>

One symptom
of this maternal-fetal conflict is its potential affect on human
health \cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. 
In addition to having an invasive
hemochorial placenta, humans possess a deeply invasive trophoblast
layer that can penetrate the myometrium of the uterus 
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
 Defects in placental growth are known to be
associated with the devastating condition of preeclampsia, which can
cause hypertension, proteinuria, and has the potential to lead to
death of the mother and/or infant \DLA{(citation)}. Although cases have been
documented in chimpanzees and gorillas \DLA{(citation)}, preeclampsia has a
far higher incidence in human pregnancies with as much as 4\% of
pregnancies being affected in the United States \DLA{(citation)}.
Chimpanzees and gorillas also have deeply invasive trophoblasts,
although the progression of invasion in great apes may be slower than
humans \cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however, large-scale expression
studies of great ape placentas have not been conducted. The increasing
affordability of next-generation sequencing technology has enabled the
large-scale study of specific tissues from a wide array of species to
address broader phylogenetic questions in the evolution of gene
expression \cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use Illumina RNA-Seq
technology to sequence and quantify the mRNA environment of six
mammalian placentas at term, including that of the \human, \bonobo (the
sister species to the chimpanzee), \dog, \cow, \elephant, and
\opossum, representing marsupials as well as three major branches of
eutherian mammals (Euarchontoglires, Laurasiatheria, Afrotheria).
These placentas also represent a wide array of placental morphologies
as well as traits such as number and size of offspring and length of
gestation \DLA{(Table 1)}. We use this data, along with existing RNA-Seq
data from mouse \cite{Wang.ea2011:surveyfornovelimprinted}, to address the major differences
between expression in marsupial and eutherian mammals at term,
identify the evolution of expression in important placental genes such
as hormones and galectins, identify novel lineage-specific expression
patterns, and quantify human specific differences in gene expression
that may be related to obstetrical disorders such as preeclampsia.

@@

Fetal tissue was collected from the placenta of six mammalian species
(\human [human], \bonobo [bonobo], \dog [domestic dog], \cow [domestic
cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \gallili [?], \carmelli [?]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in \DLA{Supplemental Table 1}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer’s instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

@@

For more complete comparisons additional sequences of placenta in
\mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
\DLA{citation}, \sheep \DLA{citation}, and \boar \DLA{citation} were
obtained from the \ac{SRA} (Supplemental \cref{tab:srr_numbers}).

@@

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (Reference of ---
in \cref{tab:ref_genome_aligned_species}) were assembled into a
reference using Trinity v2014-07-17
\cite{Haas.ea2013:Denovotranscriptsequence} and annotated using
Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
the \human or \mouse transcriptome
(\cref{tab:ref_genome_aligned_species}).
% 
All transcriptomes were then aligned to the appropriate reference
genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{Pan troglodytes} was used as
a reference for \bonobo. The separation between the two
recognized Pan species is generally perceived to be $\approx 1$ Mya
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

@@

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using topGO
\cite{topGO:2.20.0}. Gene Ontology clusters were considered
significantly enriched if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure\cite{Benjamini1995:bh_procedure} on from the
$p$ values obtained from the Fisher Exact Test were less than or equal
to 0.05. Enrichment tests were considered with two gene lists per
transcriptome: annotated genes with expression levels of the top
5\% of the frequency distribution of \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris});
the top 100 genes with the highest \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
Due to the paucity of annotated gene ontology terms for some species,
gene lists were compared using the \human Gene Ontology
using the 1:1 \human ortholog obtained

@@

All 1:1 orthologues present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via
Biomart\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the
biomaRt R package \DLA{Cite the biomaRt package}. A total of
\Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were
expressed in at least one species. Of these,
\Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  10,length(unique(human_name))]} 1:1 orthologous genes with
\ac{FPKM} $\ge 10$ were retained for future analyses of gene
expression evolution described below.
